Bayesian Deep Gaussian Processes for Correlated Functional Data: \\ A Case Study in Cosmological Power Spectra

### 2025-06-06 paper/main.tex
add appendices, change theta to psi for design, general updates to manuscript

While our inferential goal is to obtain the posterior distribution of 
$S\mid\bar{Y},\Sigma_\varepsilon$, this requires inferring latent quantity $W$.  
We first integrate over $S$ to condense our hierarchical model of Eq.~\ref{eq:dgphm} into
\begin{align}
\label{eq:likelihood}
\bar{Y} \mid W, \Sigma_\varepsilon &\sim \textrm{GP}(\mathbf{0}, \Sigma_S(W) + \Sigma_\varepsilon) \\
\label{eq:prior}
W &\sim \mathrm{GP}\left(\mathbf{0}, \Sigma_W(X)\right)
\end{align}
Details are provided in Appendix \textcolor{red}{???}. We then use elliptical slice
sampling \citep[ESS;][]{murray2010elliptical} to generate posterior samples of $W$.  
Proposals are generated from prior samples from Eq.~(\ref{eq:prior}), and acceptances 
involve the ratio of likelihoods from Eq.~\ref{eq:likelihood}.

We also embrace a fully-Bayesian treatment of unknown kernel hyperparameters 
$\{\tau^2, \theta_S, \theta_W\}$.  We prescale $X$ to $[0, 1]$
and $\bar{Y}$ to zero mean and unit variance.  On this scale, we adopt the default 
prior specifications provided in the {\tt deepgp} {\sf R}-package \citep{deepgp}.
\textcolor{red}{(@Steve - check the version number of the package in this reference.  Also,
I'm not sure what we really do with $\tau^2$.  In my package, we assign it a reference prior and
integrate it out.  I don't think we can do that here since we have the $\Sigma_\varepsilon$.)}
%\textcolor{red}{Annie, please ensure the most up-to-date 
%prior specs are shown above and here. Here is what I found from fit\$settings when using 
%deepgp v1.1.3: $\theta_S \sim G(\alpha=1.2, \beta=1)$ and $\tau_S \propto 1/\tau_S^2$.}
We integrate Metropolis-Hastings sampling of these parameters in a Gibbs framework with ESS
sampling of $W$ following the scheme of \citep{sauer2023active}.

Given burned-in posterior samples $\{W^{(t)}, \tau^{2(t)}, \theta_S^{(t)}, \theta_W^{(t)}\}$
for $t\in\mathcal{T}$, we may obtain the
posterior distribution of the true power matter spectra for a particular cosmology as
\[
S^{(t)}\mid\bar{Y}, \Sigma_\varepsilon \sim \mathcal{N}_n(m^{(t)}, C^{(t)})
\quad\textrm{where}\quad
\begin{array}{rl}
C^{(t)}&=\left(\Sigma_S(W^{(t)})^{-1}+\Sigma_\varepsilon^{-1}\right)^{-1} \\
m^{(t)}&=C^{(t)}\left(\Sigma_\varepsilon^{-1}\bar{Y}+\Sigma_S(W^{(t)})^{-1}\mu\right).
\end{array}
\] 
\textcolor{red}{(@Steve - provide details of this derivation.  Also what is $\mu$?  Are we 
going with zero here?  Add details of how we combine these things across all the $t$ samples.)}
We denote our inference procedure for the hierarchical model of Eq.~\ref{eq:dgphm} as
``DGP.HM'' given its deep Gaussian process foundation with additional hierarchical modeling (HM)
structure to incorporate multiple functional outputs.
\textcolor{red}{(Is it too late to change this?  Maybe DGP-FCO for functional correlated output?)}

@@

For simplicity, we will denote these eight parameters as $\theta \in \mathbb{R}^8$.

@@

Figure \ref{fig:mean_PCs_oneW} illustrates this decomposition with the estimated 
mean trend, the principal components contained in $B$, and one example of how the 
weight for the first PC will vary dependent on the first of $p_\theta=8$ cosmological 
parameters contained in $\theta$.

@@

% This is old: switched to SEPIA
% We assume a powered exponential kernel for the Gaussian process and employ 
% \texttt{GPfit} to perform a multi-start gradient based search for GP's hyperparameters 
% \citep{macdonald2015gpfit}. 
In alignment with \cite{heitmann2009coyote} and \cite{higdon2010estcosmo}, we use 
GPs to model the principal components' weights. For direct comparison alongside
\cite{moran2023mira}, we employ a Python implementation of this approach called 
SEPIA \citep{gattiker2020sepia}. \textcolor{red}{Please note: currently show GPfit 
predictions for Mira-Titan Data, NOT SEPIA. The CAMB results are based on SEPIA.}
That is, the $i$th vector of PC weights $w_i(\theta)$, 
$i \in \{1,\ldots,p_\eta\}$, will be modeled as a zero mean GP: 
$w_i(\theta) \sim GP(0, \sigma^2R)$, with

\begin{equation*}
    R_{ij} = \prod_{k=1}^{p_\theta}\exp\left\{-10^{\beta_k}|x_{ik}-x_{jk}|^\alpha\right\}
\end{equation*}

\textcolor{red}{We fix $\alpha=1.95$ and estimate each $\beta_k$ with $k \in \{1,...,p_\theta\}$. 
The above with $R_{ij}$ is based on GPfit; will update this to reflect SEPIA settings later.}
Using these trained GPs from the PCs, we predict the spectra for unobserved cosmologies.

@@

Both Cosmic Emu and DGP.HM are capable of modeling a true underlying functions based 
upon multiple functional realizations, but our approach utilizes DGPs within a 
hierarchical model in lieu of deep process convolutions. After obtaining the estimated 
smooth spectra from the training cosmologies, each method utilizes GPs in order to 
estimate the weights of the PCs from the estimated spectra. Below, we compare prediction 
performance between Cosmic Emu and DGP.HM for the 6 held-out cosmologies. 
\textcolor{magenta}{In addition, we can also predict over a large uniform or full 
factorial design to obtain estimated main effects for each cosmological parameter in 
$\theta$, interactions amongst the $p_\theta=8$ parameters, and decompose the overall 
variation by main effects, two-way interactions, and so on. Should this be included?}

@@

\section{Appendix}
\label{sec:apdx}

>>>>

While our inferential goal is to obtain the posterior distribution of 
$S\mid\bar{Y},\Sigma_\varepsilon$, this requires inferring latent quantity $W$.  
We first integrate over $S$ to condense our hierarchical model of Eq.~\ref{eq:dgphm} into
\begin{align}
\label{eq:likelihood}
\bar{Y} \mid W, \Sigma_\varepsilon &\sim \textrm{GP}(\boldsymbol{\mu}, \Sigma_S(W) + \Sigma_\varepsilon) \\
\label{eq:prior}
W &\sim \mathrm{GP}\left(\mathbf{0}, \Sigma_W(X)\right)
\end{align}
Details are provided in \ref{sec:apdx_int_lik}. We then use elliptical slice
sampling \citep[ESS;][]{murray2010elliptical} to generate posterior samples of $W$.  
Proposals are generated from prior samples from Eq.~(\ref{eq:prior}), and acceptances 
involve the ratio of likelihoods from Eq.~\ref{eq:likelihood}.

We also embrace a fully-Bayesian treatment of unknown kernel hyperparameters 
$\{\tau^2, \theta_S, \theta_W\}$.  We prescale $X$ to $[0, 1]$
and $\bar{Y}$ to zero mean and unit variance.  On this scale, we adopt the default 
prior specifications provided in the {\tt deepgp} {\sf R}-package \citep{deepgp}.
\textcolor{red}{(@Annie - from our email 5/29, can you comment on how we handle $\tau^2$/
cite Bobby's book/etc? Thanks!)}
%\textcolor{red}{Annie, please ensure the most up-to-date 
%prior specs are shown above and here. Here is what I found from fit\$settings when using 
%deepgp v1.1.3: $\theta_S \sim G(\alpha=1.2, \beta=1)$ and $\tau_S \propto 1/\tau_S^2$.}
We integrate Metropolis-Hastings sampling of these parameters in a Gibbs framework with ESS
sampling of $W$ following the scheme of \citep{sauer2023active}.

Given burned-in posterior samples $\{W^{(t)}, \tau^{2(t)}, \theta_S^{(t)}, \theta_W^{(t)}\}$
for $t\in\mathcal{T}$, we may obtain the
posterior distribution of the true power matter spectra for a particular cosmology as
\[
S^{(t)}\mid\bar{Y}, \Sigma_\varepsilon \sim \mathcal{N}_n(m^{(t)}, C^{(t)})
\quad\textrm{where}\quad
\begin{array}{rl}
C^{(t)}&=\left(\Sigma_S(W^{(t)})^{-1}+\Sigma_\varepsilon^{-1}\right)^{-1} \\
m^{(t)}&=C^{(t)}\left(\Sigma_\varepsilon^{-1}\bar{Y}+\Sigma_S(W^{(t)})^{-1}\boldsymbol{\mu}\right).
\end{array}
\]

These results come from Bayes' Theorem, with details of this derivation provided 
in \ref{sec:apdx_SgivenY}. In our application, we find success setting 
$\boldsymbol{\mu}=\boldsymbol{0}$. With all posterior draws $S^{(t)}$, we compute
the average of these to obtain our posterior mean for a particular cosmology.
We denote our inference procedure for the hierarchical model of Eq.~\ref{eq:dgphm} as
``DGP.HM'' given its deep Gaussian process foundation with additional hierarchical modeling (HM)
structure to incorporate multiple functional outputs.

@@

For simplicity, we will denote these eight parameters as $\psi \in \mathbb{R}^8$.

@@

Figure \ref{fig:mean_PCs_oneW} illustrates this decomposition with the estimated 
mean trend, the principal components contained in $B$, and one example of how the 
weight for the first PC will vary dependent on the first of $p_\psi=8$ cosmological 
parameters contained in $\psi$.

@@

% This is old: switched to SEPIA
% We assume a powered exponential kernel for the Gaussian process and employ 
% \texttt{GPfit} to perform a multi-start gradient based search for GP's hyperparameters 
% \citep{macdonald2015gpfit}. 
In alignment with \cite{heitmann2009coyote} and \cite{higdon2010estcosmo}, we use 
GPs to model the principal components' weights. For direct comparison alongside
\cite{moran2023mira}, we employ a Python implementation of this approach called 
SEPIA \citep{gattiker2020sepia}. \textcolor{red}{Please note: currently show GPfit 
predictions for Mira-Titan Data, NOT SEPIA. The CAMB results are based on SEPIA.}
That is, the $i$th vector of PC weights $w_i(\psi)$, 
$i \in \{1,\ldots,p_\eta\}$, will be modeled as a zero mean GP: 
$w_i(\psi) \sim GP(0, \sigma^2R)$, with

\begin{equation*}
    R_{ij} = \prod_{k=1}^{p_\psi}\exp\left\{-10^{\beta_k}|x_{ik}-x_{jk}|^\alpha\right\}
\end{equation*}

\textcolor{red}{We fix $\alpha=1.95$ and estimate each $\beta_k$ with $k \in \{1,...,p_\psi\}$. 
The above with $R_{ij}$ is based on GPfit; will update this to reflect SEPIA settings later.}
Using these trained GPs from the PCs, we predict the spectra for unobserved cosmologies.

@@

Both Cosmic Emu and DGP.HM are capable of modeling a true underlying functions based 
upon multiple functional realizations, but our approach utilizes DGPs within a 
hierarchical model in lieu of deep process convolutions. After obtaining the estimated 
smooth spectra from the training cosmologies, each method utilizes GPs in order to 
estimate the weights of the PCs from the estimated spectra. Below, we compare prediction 
performance between Cosmic Emu and DGP.HM for the 6 held-out cosmologies. 
\textcolor{magenta}{In addition, we can also predict over a large uniform or full 
factorial design to obtain estimated main effects for each cosmological parameter in 
$\psi$, interactions amongst the $p_\psi=8$ parameters, and decompose the overall 
variation by main effects, two-way interactions, and so on. Should this be included?}

@@

\appendix
\renewcommand{\thesection}{Appendix \Alph{section}}

\section{Integrated Likelihood}
\label{sec:apdx_int_lik}
Here, we establish that under the model:

$$
\begin{aligned}
Y_i|S,W &\sim N\big(S, \Sigma_\varepsilon\big), \quad i \in \{1,\dots, r\}\\
S|W &\sim N\big(\boldsymbol{\mu}, \Sigma_S(w,w')\big)\\
W &\sim N\big(\mathbf{0}, \Sigma_W(x,x')\big) \\
\end{aligned}
$$

We can obtain the integrated likelihood $p(Y_i|W)=\int p(Y_i|S,W)p(S|W) dS$, 
with the following calculations:
$$
\begin{aligned}
p(Y_i|W) &= \int p(Y_i|S,W)p(S|W)dS \\
&\propto \int\exp\left(-\frac{1}{2} \left[Y_i^T\Sigma_\varepsilon^{-1} Y_i 
- 2Y_i^T\Sigma_\varepsilon^{-1} S + S^T \Sigma_\varepsilon^{-1} S + S^T \Sigma_S^{-1} S 
-2S^T \Sigma_S^{-1} \boldsymbol{\mu} 
+ \boldsymbol{\mu}^T \Sigma_S^{-1} \boldsymbol{\mu} \right]\right)dS \\
\end{aligned}
$$

If we focus only on terms containing $S$, we have 
$$
\begin{aligned}
    \exp\Big(-\frac{1}{2} \big[S^T &(\Sigma_\varepsilon^{-1}+\Sigma_S^{-1}) S 
    -2S^T (\Sigma_S^{-1} \boldsymbol{\mu} + \Sigma_\varepsilon^{-1} Y_i) \big]\Big) \\
&= \exp\left(-\frac{1}{2} \left[S^T (\Sigma_\varepsilon^{-1}+\Sigma_S^{-1}) S 
-2S^T (\Sigma_\varepsilon^{-1}+\Sigma_S^{-1})(\Sigma_\varepsilon^{-1}
+\Sigma_S^{-1})^{-1}(\Sigma_S^{-1} \boldsymbol{\mu} + \Sigma_\varepsilon^{-1} Y_i) \right]\right)
\end{aligned}
$$

To complete the square, we add and subtract $(\Sigma_S^{-1} \boldsymbol{\mu} 
+ \Sigma_\varepsilon^{-1} Y_i)^T(\Sigma_\varepsilon^{-1}+\Sigma_S^{-1})^{-1}
(\Sigma_S^{-1} \boldsymbol{\mu} + \Sigma_\varepsilon^{-1} Y_i)$, 
which then gives the kernel for $S \sim N\left((\Sigma_\varepsilon^{-1}
+\Sigma_S^{-1})^{-1}(\Sigma_S^{-1} \boldsymbol{\mu} + \Sigma_\varepsilon^{-1} Y_i), 
(\Sigma_\varepsilon^{-1}+\Sigma_S^{-1})^{-1} \right)$. 
This removes the integral and leaves us with the following terms for $Y_i|W$:
$$
\begin{aligned}
&\exp\Big(-\frac{1}{2}\big[Y_i^T\Sigma_\varepsilon^{-1} Y + 
\boldsymbol{\mu}^T\Sigma_S^{-1} \boldsymbol{\mu} - (\Sigma_S^{-1} \boldsymbol{\mu} + 
\Sigma_\varepsilon^{-1} Y_i)^T(\Sigma_\varepsilon^{-1}+\Sigma_S^{-1})^{-1} 
(\Sigma_S^{-1} \boldsymbol{\mu} + \Sigma_\varepsilon^{-1} Y_i)\big]\Big) \\
&= \exp\Big(-\frac{1}{2}\big[Y_i^T\big(\Sigma_\varepsilon^{-1} - 
\Sigma_\varepsilon^{-1}(\Sigma_\varepsilon^{-1}+\Sigma_S^{-1})^{-1}\Sigma_\varepsilon^{-1}\big)Y_i 
-2Y_i^T \Sigma_\varepsilon^{-1}(\Sigma_\varepsilon^{-1}+\Sigma_S^{-1})^{-1}\Sigma_S^{-1} \boldsymbol{\mu} \\
&\qquad \qquad \qquad + \boldsymbol{\mu}^T (\Sigma_S^{-1} - \Sigma_S^{-1}(\Sigma_\varepsilon^{-1}
+\Sigma_S^{-1})^{-1}\Sigma_S^{-1}) \boldsymbol{\mu}\big]\Big) \\
&= \exp\Big(-\frac{1}{2}\big((Y_i-\boldsymbol{\mu})^T(\Sigma_\varepsilon+\Sigma_S)^{-1}
(Y_i-\boldsymbol{\mu}) \big)\Big)
\end{aligned}
$$

With the last equality resulting from the use of the Sherman–Morrison–Woodbury formula, 
yielding $\big(\Sigma_\varepsilon^{-1} - \Sigma_\varepsilon^{-1}(\Sigma_\varepsilon^{-1}+
\Sigma_S^{-1})^{-1}\Sigma_\varepsilon^{-1} = (\Sigma_S^{-1} - \Sigma_S^{-1}
(\Sigma_\varepsilon^{-1}+\Sigma_S^{-1})^{-1}\Sigma_S^{-1}) = (\Sigma_\varepsilon+\Sigma_S)^{-1}$. 
We also use the fact that $\Sigma_\varepsilon^{-1}(\Sigma_\varepsilon^{-1}+
\Sigma_S^{-1})^{-1}\Sigma_S^{-1} = (\Sigma_\varepsilon + \Sigma_S)^{-1}$, 
which holds since it can be shown that 
$[\Sigma_S(\Sigma_\varepsilon^{-1}+\Sigma_S^{-1})\Sigma_\varepsilon](\Sigma_\varepsilon+\Sigma_S)
=(\Sigma_\varepsilon+\Sigma_S)(\Sigma_\varepsilon+\Sigma_S)$.

Therefore, we have that $Y_i|W \sim N\big(\boldsymbol{\mu}, \Sigma_S(w,w')+\Sigma_\varepsilon(w,w')\big)$.

\section{Conditional Distribution of Power Spectrum, Given Weighted Average}
\label{sec:apdx_SgivenY}

From our model specification, we have the following:

\[
S \sim \mathcal{N}(\boldsymbol{\mu}, \Sigma_S)
\]
\[
\bar{Y} \mid S \sim \mathcal{N}(S, \Sigma_\epsilon)
\]

Then, to find the distribution of $S|\bar{Y}$, we apply Bayes' Theorem below, 
where $\propto$ indicates proportionality:
\[
P(S \mid \bar{Y}) \propto P(\bar{Y} \mid S) P(S)
\]

\[
\propto \exp\left(-\tfrac{1}{2} (\bar{Y} - S)^T \Sigma_\epsilon^{-1} (\bar{Y} - S) \right) 
\cdot \exp\left(-\tfrac{1}{2} (S - \boldsymbol{\mu})^T \Sigma_S^{-1} (S - \boldsymbol{\mu})\right)
\]

\[
\propto \exp\left(-\tfrac{1}{2} \left[ \bar{Y}^T \Sigma_\epsilon^{-1} \bar{Y} 
- 2 S^T \Sigma_\epsilon^{-1} \bar{Y} + S^T \Sigma_\epsilon^{-1} S + S^T \Sigma_S^{-1} S 
- 2 S^T \Sigma_S^{-1} \boldsymbol{\mu} + \boldsymbol{\mu}^T \Sigma_S^{-1} \boldsymbol{\mu} \right] \right)
\]

\[
\propto \exp\left( -\tfrac{1}{2} \left[ S^T (\Sigma_\epsilon^{-1} + \Sigma_S^{-1}) S 
- 2 S^T (\Sigma_\epsilon^{-1} \bar{Y} + \Sigma_S^{-1} \boldsymbol{\mu}) \right] \right)
\]

From properties of the multivariate normal distribution \citep[e.g., Equation 7.1 of][]{hoff2009first},
we can establish that the mean and covariance of $S|\bar{Y}$ will be $\boldsymbol{m}$
and $C$ respectively, where we can solve for these values with:

\[
C^{-1} = \Sigma_\epsilon^{-1} + \Sigma_S^{-1}
\]
\[
C^{-1} \boldsymbol{m} = \Sigma_\epsilon^{-1} \bar{Y} + \Sigma_S^{-1} \boldsymbol{\mu}
\]

Then,
\[
C = (\Sigma_\epsilon^{-1} + \Sigma_S^{-1})^{-1}
\]
\[
\boldsymbol{m} = \left( \Sigma_S^{-1} + \Sigma_\epsilon^{-1} \right)^{-1} 
\left( \Sigma_\epsilon^{-1} \bar{Y} + \Sigma_S^{-1} \boldsymbol{\mu} \right)
= C \cdot \left( \Sigma_\epsilon^{-1} \bar{Y} + \Sigma_S^{-1} \boldsymbol{\mu} \right)
\]

Therefore, we have the result that $S|\bar{Y} \sim \mathcal{N}(\boldsymbol{m}, C)$.

\section{Mira-Titan Predictive Plots}
\label{sec:apdx_mira_plots}
